---
title: "Assignment 09: Data Scraping"
author: "Nicole Eastman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1 Set-up
library(formatR)
library(tidyr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)

getwd()

library(tidyverse)
library(rvest)
library(lubridate)
library(ggplot2)
library(dataRetrieval)

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 Using rvest to scrape website 
theURL <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021'))

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3 Use the Selector Gadget tool to assign values to variables 
water.system.name <- theURL %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  
pswid <- theURL %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
  
ownership <- theURL %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  
max.withdrawals.mgd <- theURL %>% html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4 Create dataframe with variables, convert month to order they are scraped from the website, and mutate to create a date format column
Month <- c('01', '02', '03', '04', '05', '06', '07', '08', '09', '10', '11', '12')

df_withdrawals <- data.frame("Water_System" = water.system.name,
                               "PSWID" = pswid,
                               "Ownership" = ownership,
                                "Max_Withdrawals" = max.withdrawals.mgd,
                                "Month" = rep(1:12),
                                "Year" = rep(2021))
df_withdrawals$Month<- c("Jan", "May", "Sept", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")

df_withdrawals <- df_withdrawals %>%
  mutate(Date = my(paste(Month,"-",Year)),
         as.Date(paste("01", c("Month")), format= "%b"))
  
#5 
ggplot(df_withdrawals,aes(x=Month, y=Max_Withdrawals, group=1)) + 
  geom_line() +
  labs(x = "Month",y = "Max Daily Withdrawals")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(the_pswid, Year){
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_pswid, '&year=', Year))
  water.system.name <- the_website %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  
pswid <- the_website %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
  
ownership <- the_website %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  
max.withdrawals.mgd <- the_website %>% html_nodes('th~ td+ td') %>% html_text()

df_PSWID <- data.frame(Month = (Month),
                       Year= Year,
                       Max.withdrawals = as.numeric(max.withdrawals.mgd),
                       Date = as.Date(paste0(Month, "-", Year), format = "%M-%Y")) %>%
  mutate(Ownership = !!ownership,
         PWSID = !!pswid,
         System.name = !!water.system.name)
return(df_PSWID)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
df2015_PSWID1 <- scrape.it('03-32-010', 2015)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
df_Asheville2015<- scrape.it('01-11-010', 2015)

joined_df <- rbind(df_Asheville2015, df2015_PSWID1)

combined.plot <- ggplot(joined_df, aes(x= Month, y= Max.withdrawals, color = System.name, group = 2)) +
  geom_point() +
         labs(tite= "Durham and Asheville Maximum Water Withdrawals",
              subtitle = water.system.name,
              y= 'Max Withdrawal',
              x= 'Month')
combined.plot
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9
Year <- c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019)
pswid <- "01-10-010"
df_many_years <- map2(pswid, Year, scrape.it) %>%
bind_rows(df_many_years)

asheville.plot<- ggplot(df_many_years, aes(y= Max.withdrawals, x= as.factor(Month))) +
  geom_point() +
  geom_smooth(method="loess",se=FALSE) +
  labs(x= "Month", y= "Max Daily Withdrawals", title = "Max Daily Withdrawal from 2010-2019 in Asheville")
asheville.plot

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?