---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Nicole Eastman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1 Set-up
getwd()
library(tidyverse)
library(agricolae)
library(dplyr)
library(corrplot)

Chem.Physics <- read.csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
Chem.Physics$sampledate <- as.Date(Chem.Physics$sampledate, format = "%m/%d/%y")

#2 Theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature during July does not correlate with depth across all lakes
Ha: There is a significant correlation between lake tempuratue and depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Wrangling Data
Chem.Subset<- Chem.Physics %>%
  filter(daynum %in% c(183:213)) %>%
  select(lakename:daynum, depth, temperature_C) %>%
  na.omit()


#5 Scatter Plot
TempbyDepth <- 
  ggplot(Chem.Subset, aes(x=depth, y= temperature_C)) + 
  ylim(0,35) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x= "Lake Depth", y= "Temperature (C)")
print(TempbyDepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest anything about the linearity of this trend?

> Answer: The graph suggests that temperature in celsius generally decreases as the lake depth increases. There is a substantial distribution of points along the line of best fit which suggests that the linearity (or association) of this trend is not very strong. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 Linear Regression
Chem.Regression <- lm(Chem.Subset$temperature_C ~ Chem.Subset$depth)
summary(Chem.Regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 73.71% of the variability in temperature is explained by changes in depth and the p-value is less than 0.001 which indicates the correlation between the two variables is statistically significant. The degrees of freedom for this model is 9671 due to the large volume of recorded observations.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC
TempAIC<- lm(data= Chem.Subset, temperature_C ~ year4 + daynum + depth) 
step(TempAIC)

#10 Multiple Regression
Temp.Best.Regression <- lm(data = Chem.Subset, temperature_C ~ year4 + daynum + depth)
summary(Temp.Best.Regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests we use all three variables: year4, daynum, and depth to predict temperature in the regression. The model explains 73.97% of the observed variance which is a slight improvement from the linear regression only using depth as an explanatory variable.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 Anova and Lm tests
Chem.aov<- aov(data = Chem.Subset, temperature_C ~ lakename)
summary(Chem.aov)

Chem.lm <- lm(data = Chem.Subset, temperature_C ~ lakename)
summary(Chem.lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes because the p-value is less than 0.05 which indicates the correlation between temperature and lake is significant.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. Plotting Temp by Depth
Temp.Lake.Plot<- ggplot(Chem.Subset, aes(x= depth, y = temperature_C, col= lakename)) +
  ylim(0,35) +
  geom_point(alpha= 0.5) +
  geom_smooth(method = 'lm') +
  labs(x= "Lake Depth", y= "Temperature (C)")
print(Temp.Lake.Plot)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Tukey HSD Test
Chem.HSD <- TukeyHSD(Chem.aov)
Chem.HSD
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake' and Paul Lake's mean temperature is statistically the same as Peter Lake's mean temperature. Central Lake is statistically distinct from the others.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: The HSD test and the two-sample T-test will identify whether Peter Lake and Paul Lake have distinct mean temperatures as well. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18 Crampton and Ward Lake Significance
Chem.Crampton.Ward<- Chem.Subset %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

Chem.TTest<- t.test(Chem.Crampton.Ward$temperature_C ~ Chem.Crampton.Ward$lakename)
Chem.TTest

Chem.Two.Sample <- lm(Chem.Crampton.Ward$temperature_C ~ Chem.Crampton.Ward$lakename)
Chem.Two.Sample

```

>Answer: The test gives a mean of 15.48 for Crampton Lake and 14.46 for Ward Lake; therefore, the means are similar but not equal. The Tukey test indicated similar means between the two lakes but not quite equal as well.
