---
title: "Assignment 2: Coding Basics"
author: "Nicole Eastman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. Use the sequence function using the format: (<1 number in sequence>, <last number in sequence>, by= <interval between numbers>), assign the sequence a name
seq(1,100, by=4)
sequence1<- seq(1,100, by=4)

#2. Use the mean() and the median() functions to solve and assign them a name
mean1<- mean(sequence1)
mean1

median1<- median(sequence1)
median1

#3. ask R if the mean is greater than, less than, or equal to
mean1>median1
mean1<median1
mean1==median1

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
names<- c("Nicole","Andrew","Judy","Chris")
names
scores<- c(98,97,96,95)
scores
passed<-c(TRUE, TRUE, TRUE, TRUE)
passed

testresults<- data.frame(names,scores,passed)
testresults

colnames(testresults)<- c("Student", "Test Score", "Passing")
testresults

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: By using a data frame you can have different classes of data such as a numeric column and a column containing names. Matrices only contain numerical values and they do not have titles for the rows and columns. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

ifelse(scores>=50, print("True"), print("False"))

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The ifelse function worked because the if, else option gives an error stating the length of the result does not match the length of the conditional. 
